% =====================================================================
% Short, dated predictions note for arXiv — the public, third-party
% time-stamp of our prospective predictions (posted before MD2/MD3).
% Build: pdflatex arxiv_predictions.tex
% =====================================================================
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,hyperref}
\hypersetup{colorlinks=true, allcolors=blue!60!black}
\title{\bfseries Game's gone: pre-registered manipulability in the 2026 World Cup}
\author{Francisco A. Zabala \\ \small California Institute of Technology \\ \small ORCID: 0009-0003-4867-6886}
\date{\today}
\begin{document}
\maketitle

\begin{abstract}
\noindent
Before Matchday 2 and Matchday 3 of the 2026 FIFA World Cup are played, we state the matches our
model flags as \emph{manipulable} --- final-group-stage decisions in which a team's
advancement-optimal action is not to maximize its chance of winning. The purpose of this note is to
place those predictions, and the rule for scoring them, on the public record with a verifiable
timestamp. The expanded 48-team format advances the eight best third-placed teams, which we find
raises final-matchday manipulability by roughly $1.7\times$ relative to the 32-team format and,
uniquely, creates \emph{cross-group} manipulability that simultaneous kickoffs cannot remove. We name
the five Matchday-3 matches most likely to contain a manipulable team and commit, in advance, to how
we will evaluate them after the group stage concludes.
\end{abstract}

\section{Setup}
A team's final group match is a \emph{manipulable state} when the action maximizing its expected
knockout progression $V_{\mathrm{adv}}$ (expected knockout rounds won; $0$ if eliminated) differs from
the action maximizing its probability of winning that match. We simulate the full tournament from the
official 2026 draw under two independent team-strength models --- a time-varying Elo model and a
hierarchical Bayesian Poisson goals model --- conditioning on all $24$ real Matchday-1 results, with
Matchday 2 and Matchday 3 unplayed. For each team's Matchday-3 decision we condition its match on each
target result (win/draw/lose), simulate the remainder of the tournament through the official Round-of-32
bracket, and compare advancement values. A flagged incentive is \emph{cross-group} when the
advancement-optimal action qualifies the team via the best-third pool, i.e.\ depends on other groups.

\section{What the simulation finds}
Final-matchday manipulability is markedly higher under 48 teams: the expansion multiplier is
$\rho_{48}/\rho_{32}=1.69$ (95\% CI $[1.59,1.80]$).\footnote{The multiplier and its decomposition are
estimated in a matched-strength simulation; the named Matchday-3 predictions below additionally use the
official 2026 Round-of-32 bracket. Full methods, robustness, and code are in the repository at the
commit below.} A factorial decomposition attributes this almost entirely to the best-third rule
($\rho_A/\rho_B=1.72$, $[1.63,1.81]$, holding field, group count, and draw fixed), not the larger field
or group count ($\rho_B/\rho_C=0.98$, $[0.92,1.06]$). Cross-group manipulability is present in $49\%$ of
manipulable 48-team states and is identically $0\%$ under 32 teams. A restricted theorem explains why:
group qualification is monotone in own performance (no threshold-tanking), so manipulability is
path-valued, and the best-third rule injects a cross-group advancement channel that within-group
simultaneity cannot neutralize.

\section{The predictions (frozen before MD2/MD3)}
Cross-model agreement on per-match $P(\text{manipulable})$ is Spearman $\rho_s = 0.964$. The
\textbf{registered robust set} (matches with $P(\text{manipulable})\ge 0.50$ under \emph{both} models):

\begin{center}
\begin{tabular}{llcc}
\toprule
Group & Match & $P(\text{manip})$ elo/pois & $P(\text{cross})$ elo/pois \\
\midrule
C & Brazil vs Scotland        & 56\% / 78\% & 29\% / 36\% \\
B & Canada vs Switzerland     & 57\% / 74\% & 28\% / 41\% \\
A & Czech Republic vs Mexico  & 55\% / 67\% & 16\% / 22\% \\
J & Algeria vs Austria        & 54\% / 65\% & \textbf{39\% / 51\%} \\
F & Netherlands vs Sweden     & 56\% / 58\% & 35\% / 40\% \\
\bottomrule
\end{tabular}
\end{center}

\noindent
\textbf{Registered cross-group teams} ($P(\text{cross\_group})\ge 0.20$ under both models): Algeria,
Austria (Group~J), Sweden, Netherlands, Japan (Group~F). Algeria--Austria is the strongest cross-group
match; Group~F is the cross-group hotspot.

\section{How we will score these (committed in advance)}
\textbf{Primary (behavior-independent).} H1: the multiplier's 95\% CI excludes $1$. H2: the cross-group
share is positive under 48 teams and zero under 32. Both are decided by the frozen computation and the
realized final standings; no team behavior is required.

\textbf{Secondary (behavioral, explicitly confoundable).} For the robust set we will compare each
implicated team's Matchday-3 passivity (shot rate, pressing intensity, in-play win-probability
volatility, standardized against a strength-matched historical baseline) using a pre-specified composite
index with Benjamini--Hochberg correction. By our pre-set power rule the expected number of realized
robust manipulable matches is $E=2.76<3$, so this endpoint is reported as descriptive/illustrative only;
the behavior-independent endpoints carry the test.

\section{Integrity}
The model, predictions, and scoring rule are fixed at repository commit
\texttt{ea3086323bb74afa21e96600e177d854ad5cac7a} (tag \texttt{freeze-2026-06-18}). The two frozen
prediction artifacts have SHA-256 digests
\texttt{b831d36027ebba4287c2dc7544d1f86cfe64228f4b7842a8a4f8f68c51e439c1} (Elo) and
\texttt{50783eb83565e674e428a04d4c8979f16f3fa3c1dcece54fedfafe8ba85faebd} (Poisson). The arXiv
submission date of this note is the public timestamp establishing that these predictions predate
Matchday 2 and Matchday 3.

\end{document}
