% =====================================================================
% World Cup 2026 manuscript — iteration skeleton.
% For fast iteration we use the standard `article` class with a
% Nature-like structure (summary paragraph, ~2500-word main text,
% <=30 refs). At the REVISION stage Nature/Springer requires the
% official Springer Nature class (sn-article.cls / sn-jnl). Swap then.
% Build:  latexmk -pdf main.tex   (or: pdflatex; bibtex; pdflatex x2)
% =====================================================================
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{authblk}
\usepackage{lineno}        % line numbers for review
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, allcolors=blue!60!black}

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}

\title{\bfseries Game's gone: pre-registered manipulability in the 2026 World Cup\\
\large\normalfont\itshape the best-third rule, not the larger field, increases group-stage manipulability}

\author[1]{Francisco A. Zabala}
\affil[1]{California Institute of Technology}
% \todo{add co-authors if any}

\date{\today}

\begin{document}
\maketitle
\linenumbers

% ---- Summary paragraph (Nature: ~200 words, fully referenced) --------
\begin{abstract}
\noindent
Selection systems that admit more entrants are widely assumed to be fairer, but breadth can
silently cost \emph{strategyproofness} --- the property that a competitor never gains by
underperforming. We study the 2026 expansion of the FIFA World Cup from 32 to 48 teams, whose
group stage advances, besides the top two of each group, the eight best third-placed teams. We
define a \emph{manipulable match-state} as one in which a team's advancement-optimal action is not
to maximise its chance of winning, and estimate its prevalence by Monte-Carlo simulation of the
full tournament under calibrated team-strength models. Final-matchday manipulability is markedly
higher under the 48-team format: the expansion multiplier is
$\rho_{48}/\rho_{32}=1.69$ (95\% CI $[1.59,1.80]$). A factorial decomposition shows this is caused
almost entirely by the best-third rule ($1.72\times$, holding field, groups and draw fixed), not by
the larger field or group count ($0.98\times$, indistinguishable from no effect). The rule uniquely
creates \emph{cross-group} manipulability --- incentives that route through other groups' results and
therefore survive the simultaneous final kickoffs FIFA introduced after the 1982 ``Disgrace of
Gij\'on'' --- present in $49\%$ of manipulable 48-team states and $0\%$ under 32 teams. A restricted
theorem explains the mechanism: group qualification is monotone in own performance (no
threshold-tanking), so manipulability is path-valued, and the best-third rule injects a cross-group
advancement channel that within-group simultaneity cannot neutralise. We pre-register, before the
group finale is played, the specific 2026 matches our model flags as manipulable. The price of
expansion is a quantifiable, mechanistically identified increase in manipulability.
\end{abstract}

% =====================================================================
\section*{Introduction}
Whether a competition is \emph{strategyproof} --- whether a participant can ever gain by not
trying to win --- is a mechanism-design property with a long lineage in social choice, where the
Gibbard--Satterthwaite theorem \cite{gibbard1973,satterthwaite1975} shows that non-trivial choice
rules are generically manipulable. Tournament design is the sporting instance of the same question
\cite{csato2021tournament,kendall2017sportsrules}, and its failures are concrete: at the 1982 World
Cup in Gij\'on, West Germany and Austria played out a result that eliminated Algeria, after which
FIFA mandated that a group's final two matches kick off \emph{simultaneously} so that neither team
can play to a known target. Simultaneity removes manipulation that depends on a contemporaneously
played match \emph{within a team's own group}.

The 2026 expansion to 48 teams changes the advancement rule, not merely its scale: twelve groups of
four send their top two plus the eight best third-placed teams to a Round of 32. A third-placed
team's fate is decided by comparison with teams in \emph{other} groups --- a cross-group dependency
that within-group simultaneity cannot conceal. Existing analyses of the expansion treat
competitiveness and balance \cite{csato2025imbalanced} or forecast outcomes; the strategyproofness
of the new format, and a clean attribution of any change to the rule rather than the larger field,
have not been established. Incentive incompatibility of qualification more broadly is known
\cite{dagaev2018winning,pauly2014roundrobin,vong2017strategic,guyon2015rethinking}, but as
existence, not as a measured, mechanistically decomposed, and prospectively tested quantity.

We make three contributions. First, we define and measure the prevalence of manipulable
final-matchday states under both formats. Second, we causally isolate the best-third rule by a
factorial that holds the field, group structure and draw fixed, and prove a restricted theorem that
explains why the rule, and not the expansion per se, creates manipulability that simultaneity cannot
remove. Third, we pre-register the specific 2026 matches our model flags, turning the tournament
into a prospective natural experiment rather than a retrospective fit.

\section*{Results}
\subsection*{Manipulable match-states and the expansion multiplier}
We simulate each format's full bracket under two independent team-strength models (Methods), and at
every team's final-group-match decision compare the action that maximises its probability of winning
the match with the action that maximises its advancement value $V_{\mathrm{adv}}$ (expected knockout
rounds won). A state is manipulable when these differ (Methods). Across the official 2026 draw, the
final-matchday manipulability rate is $\rho_{48}=0.27$ versus $\rho_{32}=0.16$, an expansion
multiplier of $\rho_{48}/\rho_{32}=1.69$ (95\% CI $[1.59,1.80]$; cluster bootstrap over Monte-Carlo
snapshots). The interval excludes 1 across model families, manipulability margins, and Monte-Carlo
budgets (Robustness). \todo{Figure 1: $\rho$ by format with CIs; convergence-vs-snapshots inset.}

\subsection*{The cause is the rule, not the field}
The multiplier conflates several simultaneous changes (field size, group count, advancement rule,
draw). We isolate the best-third rule with a three-arm factorial: (A) the real 48-team format; (B)
the identical 48-team field, groups and draw but with only the top two advancing; and (C) the 32-team
baseline. The multiplier decomposes multiplicatively as $\rho_A/\rho_C=(\rho_A/\rho_B)(\rho_B/\rho_C)$.
The best-third-rule term is $\rho_A/\rho_B=1.72$ (95\% CI $[1.63,1.81]$), while the field-plus-group
term is $\rho_B/\rho_C=0.98$ (95\% CI $[0.92,1.06]$), statistically indistinguishable from no effect.
Cross-group manipulability --- states whose advancement-optimal action routes through the best-third
pool --- accounts for $49\%$ of manipulable states under the rule and is identically $0\%$ without it
(arms B and C). Expanding the field without the wildcard rule produces essentially no additional, and
no cross-group, manipulability. \todo{Figure 2: decomposition bars with CIs.}

\subsection*{A restricted theorem}
Two propositions explain the simulation (Methods; full statements and proofs in the Supplementary
Theory). \emph{Proposition 1:} holding others' results fixed, the set of a team's own results for
which it advances is upward-closed in $\{\mathrm{LOSE}\prec\mathrm{DRAW}\prec\mathrm{WIN}\}$ in both
formats; hence winning maximises the probability of advancing and there is no pure-qualification
manipulation. All manipulability is therefore path/seeding-valued --- consistent with the simulated
qualification-objective rate being $\approx 0$ in both formats. \emph{Proposition 2:} the best-third
rule creates states in which a team's advancement-optimal action is not to win and the deciding
dependency is cross-group --- it still qualifies as a best third (a comparison across groups) and a
lower finish routes it, via the fixed Round-of-32 assignment, to a weaker path. Because the deciding
quantities lie outside the team's group, within-group simultaneity cannot remove the incentive. No
state of this type exists under 32 teams, where third place is elimination. We do not claim the
32-team format is manipulation-free --- seeding manipulation exists at every breadth --- only that
the best-third rule adds a cross-group channel; the magnitude is the empirical decomposition above.

\subsection*{Robustness}
The multiplier is $1.64$--$1.74$ across the two strength-model families (named-prediction rank
correlation $\rho_s=0.90$), three manipulability margins (rank correlation $\approx0.98$), and is
stable once the Monte-Carlo budget exceeds $\approx 80$ snapshots, settling at $1.64\,[1.58,1.71]$ at
400 snapshots; confidence intervals use a cluster bootstrap that resamples whole snapshots, the
correct unit given within-snapshot correlation. Under alternative advancement values --- probability
of qualifying, and probability of winning the title --- the qualification objective yields no
manipulability in either format (Proposition 1) and the title objective is too sparse to estimate
reliably; the canonical rounds-won objective is reported throughout.

\subsection*{Pre-registered live test on the 2026 group finale}
Before the group finale is played we register the matches our model flags as containing a manipulable
team, ranked by probability and labelled by whether the incentive is cross-group (Methods; the frozen
artifact and protocol are version-controlled and time-stamped). The behaviour-independent core
(H1: multiplier $>1$; H2: cross-group share $>0$ only under 48 teams) is decided by the frozen
computation. A secondary, explicitly confoundable analysis compares the implicated teams' observed
passivity against a strength-matched historical baseline. \todo{Insert frozen named-match table (R11)
and, after the finale, the scored outcome.}

\section*{Discussion}
Our results separate two things usually conflated in debates over expanding a competition: admitting
more entrants, and the rule used to admit them. The manipulability cost of the 2026 World Cup is not
a price of size; it is a price of the specific best-third wildcard, which makes advancement depend on
other groups in a way the post-Gij\'on simultaneity fix cannot reach. The same logic applies to any
selection system that admits a marginal tier by cross-comparing otherwise-separate competitions ---
wild-card play-offs, cross-panel grant or hiring quotas, multi-region qualification --- where the
remedy is structural (synchronise or sever the cross-comparison), not exhortatory. The principal
limitation is that our behavioural test of manipulation is confoundable by rotation, fatigue and
tactics; we therefore make the behaviour-independent manipulable-state count the primary claim and
pre-register it. Remaining modelling extensions --- a finer action space capturing within-win
scoreline targeting, an equilibrium (rather than best-response) treatment of the simultaneous finale,
and goal-model calibration with the Dixon--Coles low-score correction \cite{dixoncoles1997} --- are
expected to sharpen, not overturn, the identified mechanism.

\section*{Methods}
\paragraph{Data and strength models.} We fit two independent models to historical international
results: a time-varying Elo with an Elo-to-Poisson scoreline sampler, and a hierarchical Bayesian
Poisson (attack/defence with home advantage, non-centred, sampled by NUTS with convergence verified
for the 48 finalists). Probabilities are assessed by proper scores (Brier, log-loss, ranked
probability score) \cite{gneiting2007scoring} rather than accuracy. The current Poisson is
independent (no Dixon--Coles low-score correlation); calibration and a corrected refit are reported
as robustness.
\paragraph{Simulator and bracket.} A Monte-Carlo engine plays the group stage under FIFA ranking
criteria (points, goal difference, goals scored, head-to-head) and the official 2026
Round-of-32$\rightarrow$final bracket, including the candidate-set assignment of the eight best
third-placed teams to bracket slots. Format and draw are the swapped variables.
\paragraph{Manipulable states and $V_{\mathrm{adv}}$.} For a team's final-group match we condition
the scoreline on each target result $\{\mathrm{WIN},\mathrm{DRAW},\mathrm{LOSE}\}$, simulate the
remainder against the fixed strength model (a decision-theoretic best response, not an equilibrium),
and set $V_{\mathrm{adv}}$ to expected knockout rounds won (zero if eliminated; this subsumes
advancement probability and bracket-path strength). A state is manipulable when the
$V_{\mathrm{adv}}$-maximising action is not WIN by more than a noise margin; it is cross-group when
the advancement-optimal action qualifies the team via the best-third pool. Estimands, the cluster
bootstrap, the convergence-justified budget, and the full pre-registration protocol are given in the
Supplementary Information.
\paragraph{Pre-registration.} The model, predictions and scoring rule were fixed and time-stamped
before the group finale; the protocol, frozen prediction artifact and analysis code are public.

% ---- Back matter ----
\paragraph{Data availability.} Historical results and the official 2026 draw/schedule are in the
repository; \todo{dataset DOI}.
\paragraph{Code availability.} \url{https://github.com/datasith/world-cup-2026-research}
\paragraph{Prospective predictions.} The named Matchday-3 predictions and the scoring rule were
fixed and publicly time-stamped before Matchday~2 and Matchday~3 via a dated arXiv note
(\todo{arXiv id}) and a signed repository tag (\texttt{freeze-2026-06-18}, commit
\texttt{ea30863}); the frozen prediction artifacts are archived at Zenodo (\todo{Zenodo DOI})
with the SHA-256 digests recorded in that note. Protocol: \texttt{PREREGISTRATION.md}.

\bibliographystyle{naturemag}
\bibliography{references}

\end{document}
